DRAFT: Social Learning and Shared Belief: The Importance of Bias and Trust for Information Diffusion

### 2014-07-07 paper/paper.tex
fix paper

We draw heavily from Gentzkow and Shapiro's 2006 paper \emph{Media Bias and Reputation}\footnote{Gentzkow and Shapiro.  Media Bias and Reputation. 2006}.  Gentzkow and Shapiro develop a model where media firms bias their reporting to align with the beliefs of their market to maximize profits.  We generalize this model to a network of media firms, where each firm learns from the actions of other firms in the network.

To estimate media bias we consider a number of different measures drawn from several different sources.  Groseclose and Milyo develop a measure that looks at the citations made by different media firms and how these coincide with the citations made by members of Congress\footnote{Groseclose and Milyo. A Measure of Media Bias. 2005.}.  Similarlly, Gentzkow and Shapiro develop a measure of media slant that uses the similarity between words used by different media outlets and members of Congress\footnote{Gentzkow and Shapiro.  What Drives Media Slant?  Evidence from U.S. Daily Newspaper. 2006.}.  Finally, Chiang and Knight develop a model of media bias that looks at endorsements made by media firms\footnote{Chiang and Knight.  Media Bias and Influence: Evidence from Newspaper Endorsements. 2011.}.

@@

The key work we draw from in the social learning literature is Acemoglu et. al's 2011 paper \emph{Bayesian Learning in Social Networks}\footnote{Acemoglu, Dahleh, Lobel and Ozdaglar.  Bayesian Learning in Social Networks. 2011.}.  They develop a model of Bayesian social learning the is built around a stochastic network similar to the one we develop here.  They derive a number of general conclusions that are significant to this type of model. Acemoglu et. al's work on social learning in endogenous social networks is also significant as it mirrors the network structure we develop here\footnote{Acemoglu, Bimpikis and Ozdaglar.  Dynamic of Information Exchange in Endogenous Social Networks.  2013.}.  Non Bayesian approaches have also been considered in the literature including Jadbabaie et. al's work\footnote{Jadbabaie, Molavi, Sandroni and Tahbaz-Salehi.  Non-Bayesian Social Learning.  2012.}.

There is a large literature on more general information diffusion.  Young provides a good overview of how the social learning methods popular in economics compare to those used in other fields\footnote{Young.  Innovation Diffusion in Heterogeneous Populations: Contagion, Social Influence and Social Learning.  2014.}.  Golub and Jackson study how homophily affects diffusion and learning in networks.  Their results are directly relevant to the work done here\footnote{Golub and Jackson.  How Homophily Affects Learning and Diffusion in Networks. 2009.}.
 Yang and Leskovec develop a general model of information diffusion in implicit networks.  This is significant here because the network structure for our model is implicit\footnote{Yang and Leskovec.  Modeling Information Diffusion in Implicit Networks.  2010.}.

The method we use to estimate our model has its roots in the topic modeling literature, in particular the idea of LDA, first proposed by Blei, Ng and Jordan\footnote{Blei, Ng, Jordan. Latent Dirichlet Allocation. 2003.}.  LDA is essentially a hierarchical model where it is assumed that each word in a document is drawn from some topic distribution that is drawn for each document.  Using this hierarchical model Blei, Ng and Jordan proposed a method to estimate the underlying topic distribution for a set of documents.  Since their seminal paper the literature has exploded with examples of topic models.  Blei provides a good survey of the methodology\footnote{Blei.  Probabilitic Topic Models. 2012}.

Two key variants that are important here are dynamic topic models and author topic models.  A combination of these ideas form the basis for the estimation procedure used in this paper.  Blei
and Lafferty introduce the idea of dynamic topic models to represent how topic may change over time\footnote{Blei and Lafferty.  Dynamic Topic Models. 2006}.  They assume that the underlying topic proportions are drawn from some multivariate normal distribution centered around the prior topic distribution.  While we don't use the same implementation here, the idea a dynamic topic is essential to our work as our estimates for the trust edges represent how much a news story influences the underlying topic distribution for other media firms.  Rosen-Zvi et. al present another important variant in the idea of author topic models\footnote{Rosen-Zvi, Griffiths, Steyvers, Smyth.  The Author-Topic Model for Authors and Documents. 2004.}.  Their model assumes that each author has some underlying preference for certain topics that can be used to cluster documents.  While we don't directly implement this procedure the idea of a biased view of topics tied to each author is important to our model.

A number of studies have used similar methods to investigate similar datasets to ours.  Hong and Davison investigate and author topic model in the Twitter context to determine how Twitter users cluster and share information\footnote{Hong and Davison. Empirical Study of Topic Modeling in Twitter. 2010.}.  Grant et. al develop an online approach to estimating topics in the Twitter search context\footnote{Grant, George, Jenneisch and Wilson. Online Topic Modeling for Real-time Twitter Search. 2011.}.  Zhao et. al use a topic model to compare Twitter media to traditional media sources\footnote{Zhao, Jiang, Weng, He, Lim, Yan and Li. 2011.}.  Finally, Blei and Lafferty use a correlated topic model to estimate correlation within the scientific community using an academic article data set similar to the one used here\footnote{Blei and Lafferty. A Correlated Topic Model of Science. 2007.}.

@@

The roots of our model lie the Gentzkow and Shapiro paper \emph{Media Bias and Reputation}.  As with the Gentzkow and Shapiro model, we assume that there are two states of the world

@@

Gentzkow and Shapiro show that, in the case where a consumer receives no exogenous signal about the state of the world, their posterior beliefs about the quality of their monopolist news provider is a strictly increasing function of the likelihood ratio

@@

This suggests that the higher $\theta_i^c$, the more likely that a consumer will believe the $i$th firm is of high quality.  In equilibrium they show that the expected gain for publishing $\hat{r}$ is

@@

In the above, $\lambda_i(\hat{s},0)$ represents the consumers posterior belief about the quality of the $i$th firm in the case of no feedback and $\lambda_i(\hat{s},S)$ represents the consumers posterior belief about the quality of the $i$th firm in the case of feedback.  $\theta_i(s)$ is the $i$th firms posterior belief about the state of the world after receiving their signal $s_i$. Gentzkow and Shapiro show that under equilibrium conditions it is in a normal media firms interests to skew their reporting to align with the consumers priors about the state of the world.

\subsection{Inter-Firm Trust}

We now consider the situations in which firms can learn from each other.  We assume that in equilibrium the $i$th firm generally publish articles that align with their consumers priors, $\theta_i^c$ and that the firms priors should reflect this, so $\theta_i \approx \theta_i^c$.  $\theta_i$ can then be thought of as the proportion of news stories published by the $i$th firm that correspond to the $R$ state of the world.  So then over $T$ periods we would expect $\theta_iT$ reports to be $\hat{r}$.  Let's now turn to the $j$th firm's beliefs about the quality of the $i$th firm over $T$ periods.  In each period, the $j$th firms beliefs should be

@@

where $\hat{\textbf{s}}_{it}$ is a vector of the reports made by the $i$th firm.  Assuming that the beliefs follow the monotone likelihood ratio property we can show that

@@

 Unfortunately, this does not work in the case considered above because each firm can observe the reports of their fellow firms, so it is possible you may observe duplicate information.  To account for this we need to account for how much firms trust each other and how likely they are to borrow information from other firms.  Fortunately, you have a measure of how much likely it is that each firm will incorporate the reports of another firm into their report, $p_j(\text{high}_i|\hat{\textbf{s}}_{it})$.  We could modify the above posterior to account for this, let $\gamma_ij$ represent $p_j(\text{high}_i|\hat{\textbf{s}}_{it})$
 x
 \[p(R|\hat{\textbf{s}}) \propto (1-\gamma_{12})(1-\gamma_{13})...(1-\gamma_{1N})p(\hat{s}_1|R)...\]

@@

In this case the more likely either firm trust the other ($\gamma_{12}$ is higher or$\gamma_{21}$ is higher), the lower $p(R|\hat{\textbf{s}})$.  That is that if two firms share a similar belief about the state of the world and report the same signal, this says less than if two firms that have different beliefs report the same signal.

@@

Next we consider the possibility for border agents.  The intuition behind the idea for border agents is that they are agents who have a high enough $\gamma$ value for two different agents with rather low $\gamma$ values that they allow the communication of ideas between groups who otherwise would not communicate.  More formally, consider a network with 3 agents, $i$, $j$, and $k$.  If

@@

the $i$th agent is a border agent.  Border agents are important because they mean that given a lag two groups that otherwise would not receive the same information can receive the same information.

@@

Next we consider the possibility that there are risk averse media firms.  These are firms that are willing to pay a premium to avoid poor quality information.  This premium may involve waiting longer to verify a report or only trusting higher quality firms.  We say that these firms are more likely to trust firms that have a higher quality level controlling for bias than less risk averse firms.  To estimate who the risk averse firms are we could assume the following relationship between trust, bias and bias-free quality

@@

\[\theta_m \sim \text{Dir}(\alpha)\]

\[z_n \sim Mult(\theta_m)\]

\[w_n \sim p(w_n | z_n, \beta) = Mult(\beta_z)\]

That is, $\theta$ is drawn from a Dirichlet distribution for each document.  Then $z_n$ is drawn for each word in the document, and the word is from a Multinomial disttribution contional on the specific $z_n$ drawn and $\beta$.  $\beta$ is a probability matrix where $\beta_{ij} = p(w^j = 1|z^i =1)$.  What makes LDA particularly interesting among the class of clustering models is that topic parameter is sampled for each individual word, allowing for a mixture of topics within a document.  Ultimately LDA allows for the estimation of $\theta$ and $\textbf{z}$ which estimate the distribution of topics within a collection of documents.

@@

@@

\[w_{tn} \sim Mult(\pi(\beta_{t, z}))\]

This generation procedure allows the distribution of topics to change over time, accounting for shifts in the content of documents.  Such a model is important for the dataset we wish to use where topics are changing regularly.  We use a variant of the Dynamic Topic Model that allows for the distribution of $\alpha_{it}$ to be impacted by $\alpha_{-it-1}$.  That is we assume that the underlying topic distribution is ultimately effected by the topic distributions for other agents in a network of agents.  More formally we say that,

\[\alpha_{it} \sim N(\alpha_{t-1}^T \gamma, \delta^2I)\]

$\alpha$ is a vector of the $\alpha$ parameter for each of the $N$ agents, and $\gamma$ is an $N$ dimensional vector of weights for the $N$ agents.  That is, $\alpha_{it}$ is drawn from a weighted sum of the $\alpha$ values for each agent in the prior period.  $\gamma$ gives an estimate of the impact that each agent has on the topic distribution of the $i$th agent.  We use $\gamma$ to represent our trust value discussed above.  Unfortunately, this generating process does not have a conjugate form so we cannot using Gibbs sampling as used commonly with LDA.  Instead we must use a more complicated Metropolis Hastings algorithm to estimate $\gamma$.

@@

@@

@@

@@

@@

>>>>

We draw heavily from Gentzkow and Shapiro's 2006 paper \emph{Media Bias and Reputation}\footnote{Gentzkow and Shapiro.  Media Bias and Reputation. 2006}.  Gentzkow and Shapiro develop a model where media firms bias their reporting to align with the beliefs of their market to maximize profits.  We generalize this model to a network of agents, where each agent learns from the actions of other agents in the network, as well as their own exogenous signals.

To estimate media bias we consider a number of different measures.  Groseclose and Milyo develop a measure that looks at the citations made by different media firms and how these coincide with the citations made by members of Congress\footnote{Groseclose and Milyo. A Measure of Media Bias. 2005.}.  Similarlly, Gentzkow and Shapiro develop a measure of media slant that uses the similarity between words used by different media outlets and members of Congress\footnote{Gentzkow and Shapiro.  What Drives Media Slant?  Evidence from U.S. Daily Newspaper. 2006.}.  Finally, Chiang and Knight develop a model of media bias that looks at endorsements made by media firms\footnote{Chiang and Knight.  Media Bias and Influence: Evidence from Newspaper Endorsements. 2011.}.

@@

The key work we draw from in the social learning literature is Acemoglu et. al's 2011 paper \emph{Bayesian Learning in Social Networks}\footnote{Acemoglu, Dahleh, Lobel and Ozdaglar.  Bayesian Learning in Social Networks. 2011.}.  They develop a model of Bayesian social learning that is built around a stochastic network similar to the one we develop here.  They derive a number of general conclusions that are significant to this type of model. Acemoglu et. al's work on social learning in endogenous social networks is also significant as it mirrors the network structure we develop here\footnote{Acemoglu, Bimpikis and Ozdaglar.  Dynamic of Information Exchange in Endogenous Social Networks.  2013.}.  Non Bayesian approaches have also been considered in the literature including Jadbabaie et. al's work\footnote{Jadbabaie, Molavi, Sandroni and Tahbaz-Salehi.  Non-Bayesian Social Learning.  2012.}, though we don't follow that approach here.

There is a large literature on more general information diffusion.  Young provides a good overview of how the social learning methods, popular in economics, compare to those used in other fields\footnote{Young.  Innovation Diffusion in Heterogeneous Populations: Contagion, Social Influence and Social Learning.  2014.}.  Golub and Jackson study how homophily affects diffusion and learning in networks.  Their results are directly relevant to the work done here\footnote{Golub and Jackson.  How Homophily Affects Learning and Diffusion in Networks. 2009.}.
 Yang and Leskovec develop a general model of information diffusion in implicit networks.  This is significant here because the network structure for our model is implicit\footnote{Yang and Leskovec.  Modeling Information Diffusion in Implicit Networks.  2010.}.

The method we use to estimate our model has its roots in the topic modeling literature, in particular the idea of Latent Dirichlet Allocation (LDA), first proposed by Blei, Ng and Jordan\footnote{Blei, Ng, Jordan. Latent Dirichlet Allocation. 2003.}.  LDA is essentially a hierarchical model where it is assumed that each word in a document is drawn from some topic distribution that is drawn for each document.  Using this hierarchical model Blei, Ng and Jordan proposed a method to estimate the underlying topic distribution for a set of documents.  Since their seminal paper the literature has exploded with examples of topic models.  Blei provides a good survey of the methodology\footnote{Blei.  Probabilitic Topic Models. 2012}.

Two key variants that are important here are dynamic topic models and author topic models.  A combination of these ideas form the basis for the estimation procedure used in this paper.  Blei
and Lafferty introduce the idea of dynamic topic models to represent how topic may change over time\footnote{Blei and Lafferty.  Dynamic Topic Models. 2006}.  They assume that the underlying topic proportions are drawn from some multivariate normal distribution centered around the prior topic distribution.  Our approach is an application of this method.  Rosen-Zvi et. al present another important variant in the idea of author topic models\footnote{Rosen-Zvi, Griffiths, Steyvers, Smyth.  The Author-Topic Model for Authors and Documents. 2004.}.  Their model assumes that each author has some underlying preference for certain topics that can be used to cluster documents.  While we don't directly implement this procedure the idea of a biased view of topics tied to each author is important to our model.

A number of studies have used similar methods to investigate similar datasets to ours.  Hong and Davison investigate an author topic model in the Twitter context to determine how Twitter users cluster and share information\footnote{Hong and Davison. Empirical Study of Topic Modeling in Twitter. 2010.}.  Grant et. al develop an online approach to estimating topics in the Twitter search context\footnote{Grant, George, Jenneisch and Wilson. Online Topic Modeling for Real-time Twitter Search. 2011.}.  Zhao et. al use a topic model to compare Twitter media to traditional media sources\footnote{Zhao, Jiang, Weng, He, Lim, Yan and Li. 2011.}.  Finally, Blei and Lafferty use a correlated topic model to estimate correlation within the scientific community using an academic article data set similar to the one used here\footnote{Blei and Lafferty. A Correlated Topic Model of Science. 2007.}.

@@

The roots of our model lie the Gentzkow and Shapiro paper \emph{Media Bias and Reputation}.  For now we focus on the case used in their paper with a media firm and consumer.  The approach works just as well in the case of a general network of agents.  As with the Gentzkow and Shapiro model, we assume that there are two states of the world

@@

Gentzkow and Shapiro show that, in the case where a consumer receives no exogenous signal about the state of the world, their posterior beliefs about the quality of their news provider is a strictly increasing function of the likelihood ratio

@@

This suggests that the higher $\theta_i^c$, the more likely  a consumer will believe the $i$th firm is of high quality.  In equilibrium they show that the expected gain for publishing $\hat{r}$ is

@@

In the above, $\lambda_i(\hat{s},0)$ represents the consumers posterior belief about the quality of the $i$th firm in the case of no feedback and $\lambda_i(\hat{s},S)$ represents the consumers posterior belief about the quality of the $i$th firm in the case of feedback.  $\theta_i(s)$ is the $i$th firms posterior belief about the state of the world after receiving their signal $s_i$. Gentzkow and Shapiro show that under equilibrium conditions it is in a normal media firms interests to skew their reporting to align with the consumers priors about the state of the world, because their consumers are more likely to trust them in this case.
\subsection{Inter-Firm Trust}

We now consider the situations in which firms can learn from each other.  We assume that in equilibrium the $i$th firm publishes articles that align with their consumers priors, $\theta_i^c$ and that the firms priors should reflect this, so $\theta_i \approx \theta_i^c$.  $\theta_i$ can then be thought of as the proportion of news stories published by the $i$th firm that correspond to the $R$ state of the world.  So then over $T$ periods we would expect $\theta_iT$ reports to be $\hat{r}$.

Let's now turn to the $j$th firm's beliefs about the quality of the $i$th firm over $T$ periods.  In each period, the $j$th firms beliefs should be

@@

where $\hat{\textbf{s}}_{it}$ is a vector of the reports made by the $i$th firm as of period $t$.  Assuming that the beliefs follow the monotone likelihood ratio property we can show that

@@

 Unfortunately, this does not work in the case where firms can observe the reports of other firms, because it is possible you may observe duplicate information.  To account for this we need to account for how much firms trust each other and how likely they are to borrow information from other firms.  Fortunately, you have a measure of how much likely it is that each firm will incorporate the reports of another firm into their report, $p_j(\text{high}_i|\hat{\textbf{s}}_{it})$.  We could modify the above posterior to account for this, let $\gamma_ij$ represent $p_j(\text{high}_i|\hat{\textbf{s}}_{it})$

 \[p(R|\hat{\textbf{s}}) \propto (1-\gamma_{12})(1-\gamma_{13})...(1-\gamma_{1N})p(\hat{s}_1|R)...\]

@@

In this case the more likely either firm trust the other ($\gamma_{12}$ is higher or $\gamma_{21}$ is higher), the lower $p(R|\hat{\textbf{s}})$.  That is that if two firms share a similar belief about the state of the world and report the same signal, this says less than if two firms that have different beliefs report the same signal.

@@

Next we consider the possibility of border agents.  The intuition behind the idea for border agents is that they are agents who have a high enough $\gamma$ value for two different agents with rather low $\gamma$ values that they allow the communication of ideas between groups who otherwise would not communicate.  More formally, consider a network with 3 agents, $i$, $j$, and $k$.  If

@@

the $j$th agent is a border agent.  Border agents are important because they mean that given a lag two groups that otherwise would not receive the same information can receive the same information.

@@

Next we consider the possibility that there are risk averse media firms.  These are firms that are willing to pay a premium to avoid poor quality information, relative to other firms.  This premium may involve waiting longer to verify a report or only trusting higher quality firms.  We say that these firms are more likely to trust firms that have a higher quality level controlling for bias than less risk averse firms.  To estimate who the risk averse firms are we could assume the following relationship between trust, bias and bias-free quality

@@

\[\theta \sim \text{Dir}(\alpha)\]

\[z_n \sim Mult(\theta)\]

\[\phi \sim Dir(\beta)\]

\[w_n \sim  Mult(\phi_z)\]

That is, $\theta$ is drawn from a Dirichlet distribution for each document,  $\phi$ is drawn from a Dirichlet distribution for each topic.  Then $z_n$ is drawn for each word in the document, and the word is from a Multinomial disttribution for the $\phi$ for the specified topic $z_n$.  $\beta$ is a probability matrix where $\beta_{ij} = p(w^j = 1|z^i =1)$.  What makes LDA particularly interesting among the class of clustering models is that topic parameter is sampled for each individual word, allowing for a mixture of topics within a document.  Ultimately LDA allows for the estimation of $\theta$, $\phi$ and $\textbf{z}$ which estimate the distribution of topics within a collection of documents.

@@

\[\phi \sim N(\beta_t, b^2I)\]

@@

\[w_{tn} \sim Mult(\pi(\phi_z)\]

This generation procedure allows the distribution of topics to change over time, accounting for shifts in the content of documents.  Such a model is important for the dataset we wish to use where topics are changing regularly.  We use a variant of the Dynamic Topic Model that allows for the distribution of $\alpha_{it}$ to be impacted by $\text{\boldmath$\alpha$}_{t-1}$.  Where $\text{\boldmath$\alpha$}_{t-1}$ is a vector of the $\alpha_{it-1}$ values.  That is we assume that the underlying topic distribution is ultimately effected by the topic distributions for other agents in a network of agents.  More formally we say that,

\[\alpha_{it} \sim N(\text{\boldmath$\alpha$}_{t-1}^T \gamma_i, \delta^2I)\]

$\gamma_i$ is an $N$ dimensional vector of weights for the $N$ agents for the $i$th agent.  That is, $\alpha_{it}$ is drawn from a weighted sum of the $\alpha$ values for each agent in the prior period.  $\gamma_i$ gives an estimate of the impact that each agent has on the topic distribution of the $i$th agent.  We use $\gamma_i$ to represent our trust value discussed above.  Unfortunately, this generating process does not have a conjugate form so we cannot using Gibbs sampling as used commonly with LDA.  Instead we must use a more complicated Metropolis Hastings algorithm to estimate $\gamma_i$.

@@

\[\propto \exp{\left(-\frac{1}{2}[(\phi - \beta_t)^T(\phi - \beta_t) + (\beta_t - \beta_{t-1})^T(\beta_t - \beta_{t-1})]\right)}\]

@@

\[\propto \exp{\left(-\frac{1}{2}[(\theta - \alpha_{it})^T(\theta - \alpha_{it}) + (\alpha_{it} - \text{\boldmath$\alpha$}_{t-1}^T \gamma_i)^T(\alpha_{it} - \text{\boldmath$\alpha$}_{t-1}^T \gamma_i)]\right)}\]

@@

\[\propto \exp{\left(-\frac{1}{2}[(\alpha_{it} - \text{\boldmath$\alpha$}_{t-1}^T \gamma_i)^T(\alpha_{it} - \text{\boldmath$\alpha$}_{t-1}^T \gamma_i) + (\gamma_i - \mu)^T(\gamma_i - \mu)]\right)}\]

@@

\[\propto \left(\prod_{i=1}^N \prod_{k=1}^K \frac{\theta_k^{I_{z_i}=k}}{\sum_k \theta}\right) \left(\exp{-\frac{1}{2}(\theta - \alpha_{it})^T(\theta - \alpha_{it})}\right) \]

@@

\[\propto \left(\prod_{i:z_i=k}^N \prod_{v=1}^V \frac{\phi_{k, v}^{I_{w_i}=v}}{\sum_v \phi_k}\right) \left(\exp{-\frac{1}{2}(\phi_k - \beta_t)^T(\phi_k - \beta_t)}\right) \]
